---
title: "Data Science Term Project"
author: "CSORNAI, Gyula - 134706"
date: "March 4, 2017"
output: html_document
classoption: landscape
---



## Requirements

* Choose a publicly available dataset with at least 10,000 records and 10 variables.
* Do EDA, data cleaning
* Try several supervised learning methods (and several values for the parameters for each method).
* Do model selection and evaluation properly (train-test or cross validation etc.) Show various diagnostics (e.g. ROC curve for classification etc.)
* Discuss which algorithms work best on your data and possibly why.
* Bonus: Try the algorithms on the raw data without the cleaning. Are the results the same or worse?

## Data and origin

### Data description and relevance

Extraction was done by Barry Becker from the 1994 Census database. A set of reasonably clean records was extracted using the following conditions: ((AAGE>16) && (AGI>100) && (AFNLWGT>1)&& (HRSWK>0)) 
Prediction task is to determine whether a person makes over 50K a year.

### Data source

**Lichman, M. (2013).**   *UCI Machine Learning Repository* [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

**Donor** 
Ronny Kohavi and Barry Becker 
Data Mining and Visualization 
Silicon Graphics. 

```{r setup, include=FALSE}
## initial cleanup
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)

## libraries
library(pander)
library(ggplot2)

## acquiring data from source
odf<- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data", sep=",", header=FALSE)
testdf <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.test", sep=",", header=FALSE)

## The source of multiplot function
## http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/ 

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}


```

## Exploratory data analysis

* age - The age of the individual
* work_class - The type of employer the individual has
* fnlwgt - The number of people the census takers believe that observation represents
* education - The highest level of education achieved for that individual
* eduY - Highest level of education in numerical form
* mar_stat - Marital status of the individual
* occupation - The occupation of the individual
* relationship -  Contains family relationship values 
* race - descriptions of the individuals race
* sex - Biological Sex in 2 base categories
* cap_gain - Capital gains recorded
* cap_loss - Capital Losses recorded
* hrpw - Hours worked per week
* nat_ctry - Country of origin for person
* mt50K - Whether or not the person makes more than $50,000 per annum income.

### Summary statistics of the data

```{r eda_str, echo=FALSE}

## Feature engineering01 - giving the correct names

fRawHeaderBackfill <- function(f){
  names(f)[names(f)=="V1"] <- 'age'
  names(f)[names(f)=="V2"] <- 'work_class'
  names(f)[names(f)=="V3"] <- 'fnlwgt'
  names(f)[names(f)=="V4"] <- 'education'
  names(f)[names(f)=="V5"] <- 'eduY'
  names(f)[names(f)=="V6"] <- 'mar_stat'
  names(f)[names(f)=="V7"] <- 'occupation'
  names(f)[names(f)=="V8"] <- 'relationship'
  names(f)[names(f)=="V9"] <- 'race'
  names(f)[names(f)=="V10"] <- 'sex'
  names(f)[names(f)=="V11"] <- 'cap_gain'
  names(f)[names(f)=="V12"] <- 'cap_loss'
  names(f)[names(f)=="V13"] <- 'hrpw'
  names(f)[names(f)=="V14"] <- 'nat_ctry'
  names(f)[names(f)=="V15"] <- 'mt50K'
  ##str(f)
  return(f)
}

testdf2 <- subset(testdf, testdf$V1!= "|1x3 Cross validator")
testdf2$V1 <- as.numeric(testdf2$V1)

df <- fRawHeaderBackfill(odf)
tdf <- fRawHeaderBackfill(testdf2)
testdf <- NULL
testdf2<- NULL
odf <- NULL

pander(summary(df))

```

### Data cleaning

* Removing ? values
* Transforming output variable to factor
* Omitting fnlwgt: The number of people the census takers believe that observation represents.

```{r eda_cln, echo=FALSE, include=FALSE}

cdf <- subset(df, df$work_class!=" ?"&
                df$occupation!= " ?" &
                df$nat_ctry != " ?"
                )
ctdf <- subset(tdf, tdf$work_class!=" ?"&
                 tdf$occupation!= " ?" &
                 tdf$nat_ctry != " ?"
)

cdf$mt50K <- as.factor(ifelse(cdf$mt50K == " <=50K",0,1 ))

```

### Data visualization

#### Single distribution of continuous variables

```{r eda_histmpl, echo=FALSE, warning=FALSE, fig.width=10, fig.height=5}

h1 <- ggplot(cdf)+aes(x=age)+
  geom_histogram(bins=50, fill='dodgerblue3')+
  labs(
    title='Age distribution',
    subtitle='50 bins',
    x='Age groups',
    y='# of citizens'
  )+
  theme_bw()

h2 <- ggplot(cdf)+aes(x=eduY)+
  geom_histogram(bins=20, fill='dodgerblue3')+
  labs(
    title='distribution of years of education',
    subtitle='20 bins',
    x='Years spent in education',
    y='# of citizens'
  )+
  theme_bw()

h3 <- ggplot(cdf)+aes(x=cap_gain)+
  geom_histogram(bins=50, fill='dodgerblue3')+
  labs(
    title='Distribution of capital gain',
    subtitle='50 bins',
    x='Capital gain groups',
    y='# of citizens'
  )+
  theme_bw()

h3B <- ggplot(cdf)+aes(x=cap_gain)+
  geom_histogram(bins=50, fill='dodgerblue3')+
  labs(
    title='Distribution of capital log gain',
    subtitle='50 bins',
    x='Capital gain groups of\nnon-zero values - log scale ',
    y='# of citizens'
  )+
  scale_x_log10()+
  theme_bw()

h4 <- ggplot(cdf)+aes(x=cap_loss)+
  geom_histogram(bins=50, fill='dodgerblue3')+
  labs(
    title='Distribution of capital loss',
    subtitle='50 bins',
    x='Capital loss groups',
    y='# of citizens'
  )+
  theme_bw()

h4B <- ggplot(cdf)+aes(x=cap_loss)+
  geom_histogram(bins=50, fill='dodgerblue3')+
  labs(
    title='Distribution of capital log loss',
    subtitle='50 bins',
    x='Capital loss groups of\nnon-zero values - log scale',
    y='# of citizens'
  )+
  scale_x_log10()+
  theme_bw()

multiplot(h1, h2, h3, h3B, h4, h4B, cols=3)


```

#### Single distributions of factors

##### Work and occupation

```{r eda_work, echo=FALSE, warning=FALSE, fig.width=10}

fb1 <- ggplot(cdf)+aes(x=work_class,fill=work_class)+
  geom_bar() +
  labs(
    title='Distribution of different work classifications',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-60, hjust=0, size=8))


fb4 <- ggplot(cdf)+aes(x=occupation,fill=occupation)+
  geom_bar() +
  labs(
    title='Distribution of different occupations',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-60, hjust=0, size=8))

multiplot(fb1,  fb4, cols=2)


```

##### Education

```{r eda_edu, echo=FALSE, fig.width=10}

ggplot(cdf)+aes(x=education,fill=education)+
  geom_bar() +
  labs(
    title='Distribution of different levels of education',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-60, hjust=0, size=8))


```

##### Personal data

```{r eda_pers, echo=FALSE, warning=FALSE, fig.width=10, fig.height=8}

fb3 <- ggplot(cdf)+aes(x=mar_stat,fill=mar_stat)+
  geom_bar() +
  labs(
    title='Distribution of different martial status',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-45, hjust=0, size=8))

fb5 <- ggplot(cdf)+aes(x=relationship,fill=relationship)+
  geom_bar() +
  labs(
    title='Distribution of different relationships',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-45, hjust=0, size=8))

fb6 <- ggplot(cdf)+aes(x=race,fill=race)+
  geom_bar() +
  labs(
    title='Distribution of different races',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-45, hjust=0, size=8))

fb7 <- ggplot(cdf)+aes(x=sex,fill=sex)+
  geom_bar() +
  labs(
    title='Distribution of different sexes',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-45, hjust=0, size=8))

multiplot(fb3,fb5, fb6, fb7, cols=2)


```

##### Original nationality

```{r eda_nat, echo=FALSE, warning=FALSE, fig.width=10}

fb8 <- ggplot(cdf)+aes(x=nat_ctry,fill=nat_ctry)+
  geom_bar() +
  labs(
    title='Distribution of different native country',
    y='# of citizens',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))


fb8B <- ggplot(cdf)+aes(x=nat_ctry,fill=nat_ctry)+
  geom_bar() +
  scale_y_log10()+
  labs(
    title='Distribution of different native country',
    subtitle='Y axis on logarytmic scale',
    y='# of citizens\nlog',
    x=''
  )+
  theme_bw()+
  theme(legend.position="none")+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))

multiplot(fb8, fb8B, cols=2)


```

#### Wage category - the future prediction outcome

```{r eda_y, echo=FALSE, warning=FALSE, fig.width=10}

ggplot(cdf)+aes(x=mt50K,fill=mt50K)+
  geom_bar()+
  labs(
    title='Distribution of wage categories',
    subtitle='The output vairable of the modeling',
    y='# of citizens',
    x='Is the annual wage more than 50,000 USD?'
  )+
  theme_bw()+
  theme(legend.position="none")

```

### Joint distributions of different variables

#### Discrete variables

```{r eda_jdisc, echo=FALSE, warning=FALSE, fig.width=10, fig.height=12}

jd1<- ggplot(cdf)+aes(x=education, y=occupation)+
  geom_count(col='dodgerblue3')+
  labs(
    title='Joint distribution of levels of education\nand occupation',
    y='Occupation',
    x='level of education'
  )+
  theme_bw()+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))

jd2<- ggplot(cdf)+aes(x=relationship, y=occupation)+
  geom_count(col='dodgerblue3')+
  labs(
    title='Joint distribution of relationship status\nand occupation',
    y='Occupation',
    x='relationship status'
  )+
  theme_bw()+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))
  
jd3 <- ggplot(cdf)+aes(x=education, y=work_class)+
  geom_count(col='dodgerblue3')+
  labs(
    title='Joint distribution of education\nand occupation',
    y='Work class',
    x='level of education'
  )+
  theme_bw()+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))

jd4 <- ggplot(cdf)+aes(x=mar_stat, y=relationship)+
  geom_count(col='dodgerblue3')+
  labs(
    title='Joint distribution of martial status\nand relationship',
    y='Relationship',
    x='martial status'
  )+
  theme_bw()+
  theme(axis.text.x=element_text(angle=-90, hjust=0, size=8))

multiplot(jd1, jd2, jd3, jd4, cols=2)


```

#### Continuous variables

##### Dependencies between age and capital by sex


```{r eda_jcdist_sex,echo=FALSE, warning=FALSE, fig.width=10, fig.height=9}

jdc1 <- ggplot(cdf)+aes(x=age, y=log(cap_gain+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(1600,0),  colours=rainbow(4))+
  facet_wrap(~sex)+
  labs(
    title='Joint distribution of age and capital gain',
    subtitle='Faceted by sex',
    y='Capital gain\non log scale',
    x='Age'
  )+
  theme_bw()

jdc2 <- ggplot(cdf)+aes(x=age, y=log(cap_loss+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(1600,0),  colours=rainbow(4))+
  facet_wrap(~sex)+
  labs(
    title='Joint distribution of age and capital loss',
    subtitle='Faceted by sex',
    y='Capital loss\non log scale',
    x='Age'
  )+
  theme_bw()

multiplot(jdc1, jdc2, cols=1)


```

##### Dependencies between age and capital gain by work class

```{r eda_jdist_gain_workclass, echo=FALSE, warning=FALSE, fig.width=10, fig.height=9}
ggplot(cdf)+aes(x=age, y=log(cap_gain+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(1600,0),  colours=rainbow(4))+
  facet_wrap(~work_class)+
  labs(
    title='Joint distribution of age and capital gain',
    subtitle='Faceted by work class',
    y='Capital gain\non log scale',
    x='Age'
  )+
  theme_bw()

```

##### Dependencies between age and capital loss by work class

```{r eda_jdist_loss_workclass, echo=FALSE, warning=FALSE, fig.width=10, fig.height=9}
ggplot(cdf)+aes(x=age, y=log(cap_loss+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(1600,0),  colours=rainbow(4))+
  facet_wrap(~work_class)+
  labs(
    title='Joint distribution of age and capital loss',
    subtitle='Faceted by work class',
    y='Capital loss\non log scale',
    x='age'
  )+
  theme_bw()

```

##### Dependencies between age and capital loss by education

```{r eda_jdist_gain_education, echo=FALSE, warning=FALSE, fig.width=10, fig.height=9}
ggplot(cdf)+aes(x=age, y=log(cap_gain+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(900,0),  colours=rainbow(4))+
  facet_wrap(~education)+
  labs(
    title='Joint distribution of age and capital gain',
    subtitle='Faceted by education',
    y='Capital gain\non log scale',
    x='Age'
  )+
  theme_bw()



```

```{r eda_jdist_loss_education, echo=FALSE, warning=FALSE, fig.width=10, fig.height=9}

ggplot(cdf)+aes(x=age, y=log(cap_loss+1))+
  geom_bin2d()+
  scale_fill_gradientn(limits=c(900,0),  colours=rainbow(4))+
  facet_wrap(~education)+
  labs(
    title='Joint distribution of age and capital gain',
    subtitle='Faceted by education',
    y='Capital loss\non log scale',
    x='Age'
  )+
  theme_bw()

```

### Further feature engineering

I am dgoint to transform the capital gain and loss values in the following way:

$$ lncg = log(capgain+1) $$
AND

$$ lncl = log(caploss+1) $$

in order to normalize the distribution of the values. This is not a fundamental error, because the majority of the observed citizens have 0 capital gain and 0 capital loss, and whoever has some, has so much, that 1 USD per annum does not really count.

## Supervised learning methods

## Model selection and validation

## Discussion of effectiveness



